---
title: "Inference for proportions"
author: "Marianne Huebner, Michigan State University"
date: '`r Sys.Date()`'
output: word_document
---

### Example: Coin tosses
Simulate 100 coin tosses.

```{r }
heads <- rbinom(100, size = 1, prob = .5)
heads
x=sum(heads) #number of heads in the 100 coin tosses
```


There were x= `r x` heads in 100 coin tosses.

## Calculating confidence intervals for $p$

We generated the sequence of heads and tails with probability of head = 0.5.  When estimating confidence intervals for $p$, is 0.5 part of the interval? The following one sample proportion interval uses the normal approximation to the binomial distribution with or without the continuity correction.

```{r }
prop.test(x, 100)          # continuity correction TRUE by default
prop.test(x, 100, correct = FALSE)  # without continuity correction
prop.test(x, 100, conf.level=0.95)   # confidence level of 0.95 is the default
```

We cannot guarantee that 95% of all confidence intervals calculated in this way contain the true probability $p$. This is called "coverage." For small samples the Agresti-Coull interval has better coverage. Here we add 2 successes and 2 failures to each sample:
$$
\tilde{p}=\frac{x+2}{n+4}
$$

There are many others types of one sample proportion confidence intervals:

```{r}
library(binom)
binom.confint(x, 100)

```

## Confidence intervals and hypothesis tests for $p$

Save the output in an object named temp.

```{r }
temp<-prop.test(x, 100) 
```

```names(temp)``` shows us what output is generated and the names of the output objects.


Confidence interval for the proportion of heads

```{r }
temp$conf.int
```

Estimated proportion

```{r }
temp$estimate
```

P value for the hypothesis test $H_0: p=0.5$ versus $H_1: p \ne 0.5$

```{r }
temp$p.value
```


The exact test is based on binomial probabilities:

```{r }
binom.test(sum(heads),100)
```


## Coverage of confidence intervals


For a  sample size 30 and true probability 0.5, how many one proportion z intervals with 95% confidence would contain the true p=0.5?

```{r}

n=30
p=0.5
sample<-rbinom(n, 1, p)  # simulate n trials of successes and failures
x<-sum(sample)  # number of success in n trials
result <- prop.test(x, n, conf.level=0.95)  # calculate a 95% confidence interval
```
 Next check whether the true p lies within the confidence interval. If yes, return "1". If no, return "0"
 ```{r}

count=0       # 1= yes, the confidence interval contains p, 0 otherwise.
    lb <- result$conf.int[1]
    ub <- result$conf.int[2]
    if(p < ub & p > lb){
      count=1
    } else {
      count=0
    }

```

Repeat this with 1000 samples. We can do this over and over again, of we write a function.

```{r}
sim<-function(t=1000, n=100, p=0.3, clevel=0.95){
count<-rep(NA, t)

for(i in 1:t){
sample<-rbinom(n, 1, p)
x<-sum(sample)
result <- prop.test(x, n, conf.level=clevel)
    lb <- result$conf.int[1]
    ub <- result$conf.int[2]
    if(p < ub & p > lb){
    count[i]=1
    } else {
        count[i]=0
    }
}

return(100*sum(count)/t)
}

```

Let's check for 1000 intervals.

```{r}
t=1000
pct<-sim(t=1000, n=30, p=0.5, clevel=0.95)
```

Among the `r t` 95% confidence intervals based on `r t` samples there were pct = `r pct` % confidence intervals that contained the true $p$. 

Is this what you would expect?

What about 80% confidence interval coverage?

```{r}
sim(t=1000, n=30, p=0.5, clevel=0.8)
```


## Inference for comparing two proportions $p_1-p_2$

### Example Cat Bites:
Among adults with cat bites, is the proportion of women bitten higher among patients with depression compared to without depression? 
More often than women without depression?

*Reference: Hanauer DA, Ramakrishnan N, Seyfried LS (2013) Describing the Relationship between Cat Bites and Human Depression Using Data from an Electronic Health Record. PLoS ONE 8(8): e70585. doi:10.1371/journal.pone.0070585  University of Michigan, Ann Arbor.*


In the dataset extracted from electronic health records during a 3 year period
in southeastern Michigan there were 750 patients seeking medical care for cat bites. Among those 310 suffered from depression.


```{r}
bites<-matrix(c(258, 306, 310-258, 440-306 ), 2,2)
colnames(bites)<-c("female", "male")
rownames(bites)<-c("depression", "no depression")
bites
```

2-sample test for equality of proportions with continuity correction
sample 1 = depression, sample 2 = no depression
```{r}
prop.test(bites,conf.level=0.95) 
```
The results indicate that among adults with catbites, the proportion of women are higher in patients with depression than without depression.  

### Extracting confidence intervals and results from hypothesis tests from the output
```{r}
temp<-prop.test(bites,conf.level=0.95)
# names(temp) shows us what output values are available
```

Confidence interval for the difference in the proportions for $p_1$=proportion of women in sample 1 (depression), 
$p_2$=proportion of women in sample 2 (no depression)
```{r}
temp$conf.int
```

Sample proportions: 
```{r}
temp$estimate
```

P-value of the hypothesis test $H_0: p_1=p_2$ versus $H_A: p_1 \ne p_2$:
```{r}
temp$p.value
```

